Last glacial cycle glacier erosion potential in the Alps

### 2021-02-03 articles/erosion/alps-erosion.tex
Merge additions from Ian (in brackets).

    While it has long been understood that glaciers leave a specific erosional
    imprint on the landscape, observing and quantifying the long-term erosion
    and sedimentation processes has been a challenge, not least due to the
    inaccessibility of glacier beds and the slow and stochastic nature of
    glacial erosion processes \aref. \ian{Perhaps this needs more detail?}

    More recent studies have moved away from process
    understanding towards attempts at calibrating empirical erosion laws for an entire
    glacier or set of glaciers. \citet{Humphrey.Raymond.1994} correlated
    temporal variations in ice velocity and suspended sediment load during a
    surge of Variegated Glacier to establish a linear erosion law.
    \citet{Koppes.etal.2015} quantified sediment
    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
    predominant control of surface air temperature on rapid glacier erosion and
    calibrating a near-square-law from glacier velocity to erosion rate.
    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
    in the outlet stream of Franz-Josef Glacier, mapped their chemical
    composition to geologic zones of distinctive glacier speed, and also
    concluded to a near-square relationship between basal sliding and glacier
    erosion, but yielding much higher erosion rates.
    \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
    38~glaciers, showing a predominant role of glacier sliding velocity over
    climate variables, yet concluding at a sub-linear relationship to erosion.

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rate and cumulative last glacial cycle erosion
    potential are analysed. Despite aggregated uncertainties on paleoclimate,
    glacier flow and glacier erosion processes, out results provide insights
    into the diversity of the Alpine glacial erosion landscape.

@@

@@

    Modelled erosion rates, $\dot{e}$, are calculated from the modelled basal
    velocities, $u_\mathrm{b}$, using a empirical erosion power-law,
    %
    \begin{equation}
        \dot{e} = K_\mathrm{g} u_\mathrm{b}^l ,
    \end{equation}
    %
    where $K_g = 5.2\times 10^{-11}\,m^{1-l}\,a^{l-1}$ and $l = 2.34$ are
    empirical constants calibrated on quantified fjord sediment yields and
    velocities from 15~Patagonian and Antarctic Peninsula tidewater glaciers
    \citep{Koppes.etal.2015}. Alternative erosion laws are analyzed in the
    discussion part (Sect.~\ref{sec:powerlaws}).

    Our approach does not account for the feedback of glacier erosion onto
    bedrock topography and ice dynamics. Neither are other forms of erosion
    accounted for. Instead, we compute the time-integrated erosion rate,
    %
    \begin{equation}
        e =  K_\mathrm{g} \int u_\mathrm{b}^l dt,
    \end{equation}
    %
    and refer to it as the cumulative ``erosion potential''. The integration
    is numerically approximated using a time-step of 10\,a in the
    main simulation, and 100\,a in the alternative climate scenarios runs. It
    can be noted from the above formula, that with an exponent, $l$,
    higher than one, the cumulative erosion potential has a greater dependency
    on the basal ice dynamics, $u_\mathrm{b}$, than the duration of glacier
    cover.

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Temporal evolution}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Deglaciation rapid erosion}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

    The modelled, domain-integrated glacier erosion rates do not increase
    together with ice volume. On the opposite, periods of ice advance and
    maximum glaciation correspond to comparatively slow erosion
    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). While this result may
    seem paradoxical, and is modulated by the exponent, $l$, in the erosion
    power-law, it is in fact compatible with previous field-based studies that
    have evidenced an inverse relationship between air temperatures and glacier
    erosion rates \citep{Koppes.etal.2015, Cook.etal.2020}.

    Deglaciation periods, characterised by a warmer climate and above-balance
    glacier extent, yield the highest domain-integrated erosion rates of the
    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). This
    result is compatible with the field-based studies, which have evidenced
    increased glacier erosion and pro-glacial sedimentation during deglaciation
    periods \aref, including the ongoing glacier retreat \aref. Yet this
    increase has often been attributed to higher meltwater availability,
    promoting river erosion, enhanced glacier sliding, and hydrofracturing
    (plucking) \aref.

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics. The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][]{Aschwanden.etal.2012,
    Seguinot.etal.2016}. Observed feedbacks between surface meltwater and basal
    sliding, not included in the model, presumably further accelerated erosion
    during deglaciation. Rapid glacier erosion can be expected to accompany the
    ongoing 21st-century deglaciation phase.

@@

    In many places throughout the Alps, significant modelled cumulative erosion
    occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). Yet
    instantaneous erosion rates show a time-transgressive radial pattern
    (Fig.~\ref{fig:transects}), so that high-altitude rapid erosion is not
    modelled to date from expansive glaciation stages, but instead to stages of
    intermediate glacier cover (Fig.~\ref{fig:hypsogram}).

    For instance, low-elevation piedmont over-deepened troughs such as Lake
    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore, are
    here modelled to be the product of extensive, Last Glacial Maximum-like
    glaciations. Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valley of the Alps such as Lauterbrunnental
    (Fig.~\ref{fig:hypsogram}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre). While the validity of the model results at high elevation are
    limited by the 1-km resolution and the shallow-ice physics
    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
    of the modelled glacial cycle implies that the numerous Alpine mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
    temperate periods with limited ice-cover such as the present day.

@@

    Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
    erosion processes, our quantitative results are to be considered with extreme
    uncertainties. Yet to the best of our current knowledge in these respective
    areas of science, we can draw the following qualitative conclusions:
    \begin{itemize}
      \item The non-linear physics of glacier deformation and sliding, combined
        with non-linear empirical erosion laws, results in a very strong
        localization of rapid glacier erosion in regions of fast flow.
      \item Temperature-dependent creep and basal sliding provide a mechanism
        for accelerated erosion during deglaciation periods, irrespective of
        surface meltwater penetration to the glacier bed.
      \item If a non-linear glacier erosion law is used, the climate-induced
        slowdown of erosion counterbalances glacier expansion, so that
        Alpine-wide glacier erosion rates do not correlate with the ice volume.
      \item Rapid glacier erosion is restricted to low elevation during stages
        of glacier advance and maximum glaciation, but propagates up-valley to
        higher elevations during periods of glacier retreat.
      \item The diversity of the Alpine glacial erosion landscape appears to
        be the time-transgressive signature of a variety of glacial stages
        ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
    \end{itemize}

>>>>

    \ian{Furthermore, glacier erosion likely  governs, at least in part, the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through the ``glacier buzz-saw''.
      In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
      Yet, other work suggests that global weather rates remained constant over this time period \citep{Willenbring.vonBlanckenburg2010}.
      Examination of contemporary erosion rates across different climates suggest that increased temperatures lead to higher erosion rates \citep{Koppes.etal.2015,Fernandez.etal.2016}.
      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship
      between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}

    \ian{
      Observing and quantifying the long-term erosion
      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion.
      Physically-based models describe quarrying process and abrasion of bedrock by debris laden ice \citep[e.g.][]{Alley.etal.1997,Iverson.2012,Beaud.etal.2014}.
      Describing the processes physically provides a basis for understanding erosion processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into the formation of glacial landforms, such as tunnel valleys and eskers \citep{Beaud.etal.2018,Hewitt.Creyts.2019}.
      Yet, these models prove difficult to implement in many cases due to the large number of poorly constrained parameters and process, i.e. water pressure flucuations and subglacial debris concentration \citep[e.g.][]{Hallet.1979,Ugelvig.etal.2018}.}

    Instead, empirical relationships between glacier sliding and erosion can represent erosional quantities  well \citep[e.g.][]{Humphrey.Raymond.1994, Herman.etal.2015, Koppes.etal.2015} and can recreate important glacial landforms \citep[e.g.][]{Harbor.etal.1988,Macgregor.etal.2000}.
    \citet{Koppes.etal.2015} quantified sediment
    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
    predominant control of surface air temperature on rapid glacier erosion and
    calibrating a near-square-law from glacier velocity to erosion rate.
    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
    in the outlet stream of Franz-Josef Glacier, mapped their chemical
    composition to geologic zones of distinctive glacier speed, and also
    concluded to a near-square relationship between basal sliding and glacier
    erosion, but yielding much higher erosion rates.
    \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
    38~glaciers, showing a predominant role of glacier sliding velocity over
    climate variables, yet concluding at a sub-linear relationship to erosion.

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rate and cumulative last glacial cycle erosion
    potential are analysed. Despite aggregated uncertainties on paleoclimate,
    glacier flow and glacier erosion processes, out results provide insights
    into the diversity of the Alpine glacial erosion landscape. \ian{Furthermore,
      we examine the glaciologcial and conditions that cause erosion and discuss
      the implications of these conditions on understanding the relationship between climate and glacier erosion.  }

@@

    \ian{It might be worth mentioning/discussing the role of ``till dilatation''
      in the sliding relationship. There are a lot of interpretations, but
      the erosion rule is thought to be a proxy for glacier abrasion of a glacier flowing
      over a hard bed. Thus if there is a layer of till present it is arguable
      if such abrasion can occur. Given the uncertainties, I doubt this factor
      changes the story very much. However, it might be worth mentioning
      something (somewhere) about how multiple processes are lumped together in
      the sliding relationship or that till processes are uncertain due to the
      unknown distribution of till. This could ease some reviewer concerns.  }

@@

    Modelled erosion rates, $\dot{e}$, are calculated from the modelled basal
    velocities, $u_\mathrm{b}$, using a empirical erosion power-law,
    %
    \begin{equation}
        \dot{e} = K_\mathrm{g} u_\mathrm{b}^l ,
    \end{equation}
    %
    where $K_g = 5.2\times 10^{-11}\,m^{1-l}\,a^{l-1}$ and $l = 2.34$ are
    empirical constants calibrated on quantified fjord sediment yields and
    velocities from 15~Patagonian and Antarctic Peninsula tidewater glaciers
    \citep{Koppes.etal.2015}. \ian{We discuss the role of alternative values of
    $K_\mathrm{g}$ and $l$ in Sect.~\ref{sec:powerlaws}.}

    \ian{Our approach does not account for the feedback of glacier erosion onto
    bedrock topography and, then onto, ice dynamics \citep[e.g.][]{Anderson.etal.2012}.
    Neither does the approach account for downstream advection of debris \citep{Anderson.Anderson.2016}
    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}. \textbf{this last
    sentence might not be that important}}

    We compute the time-integrated erosion rate,
    %
    \begin{equation}
        e =  K_\mathrm{g} \int u_\mathrm{b}^l dt,
    \end{equation}
    %
    and refer to it as the cumulative ``erosion potential''. The integration
    is numerically approximated using a time-step of 10\,a in the
    main simulation, and 100\,a in the alternative climate scenarios runs. It
    can be noted from the above formula, that with an exponent, $l$,
    higher than one, the cumulative erosion potential has a greater dependency
    on the basal ice dynamics, $u_\mathrm{b}$, than the duration of glacier
    cover.

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\subsection{Temporal evolution}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
\ian{Instead of saying ``volumetric erosion rate (m$^3$ a$^{-1}$)'', is it wiser to use
  ``Annual erosion volume (m$^3$)'',
  Also, in many papers there are plots of erosion rate over time. For one plot,
  would it be good to plot erosion rate instead of volume (divide by area)?
  This way some comparison can be easier made with other literature.
  For instance, \citep{Stutenbecker.etal.2017} mentions a current erosion rate of $1.24$ mm a$^{-1}$.
  Such a scaling might be good for comparison... or not if things are way off
  and the message is much more about the relative change.
}

@@

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Deglaciation rapid erosion}
 \label{sect:deglac_ero}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \ian{This section title strange now, maybe \textbf{Rapid erosion during
     deglaciation}. However, what about this?  \textbf{Implications for
     understanding the relationship between erosion and climate}... I think
   it is not only the deglaciation that is interesting, but also the periods before.}

    The modelled, domain-integrated glacier erosion rates do not increase
    together with ice volume. On the opposite, periods of ice advance and
    maximum glaciation correspond to comparatively slow erosion
    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). While this result may
    seem paradoxical, and is modulated by the exponent, $l$, in the erosion
    power-law, it is in fact compatible with previous field-based studies that
    have evidenced an \ian{\textbf{do you mean inverse or direct? inverse}}
    relationship between air temperatures and glacier erosion rates
    \ian{in some cases} \citep{Koppes.etal.2015, Cook.etal.2020}.

    \ian{Cold temperatures generally reduce sliding \citep{Koppes.etal.2015} and
      if the glacier becomes cold-based then sliding, and thus erosion, will cease altogether.
      Furthermore, in cases of advance, ice surfaces gradients are lower and cold ice is
      advected downglacier, thus limiting erosion.}

    Deglaciation periods, characterised by a warmer climate and above-balance
    glacier extent, yield the highest domain-integrated erosion rates of the
    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
    \ian{\textbf{I think it would be good to have a sentence about the
        differences in glacier dynamics that cause the increase.}
      Recent work, examining a denudation records, highlights that a non-linear forcing between
      climate and glacier erosion persists \citep{Mariotti.etal.2021}. This paper
      noted an increase in glacier erosion in the southern Alps, in the Var Valley,
      roughly at $25$\,\unit{Ka} Bp following an extended period of low dunudation rates.
      The increase in erosion here, predates the increase observed in our study
      ($\sim 15$\,\unit{Ka}; Figs.~\ref{fig:evolution}). However, similar processes caused by climate warming
      likely persist. This sedimentary record \citep{Mariotti.etal.2021} and the model results
      here demonstrate that climatic conditions over-which glacier erosion occurs
      are relatively specific and likely transient. }

    \ian{I added some citations here. Increases in erosion following glacier retreat
      seems to happen. But there are good critisims of the sedimentary record, and the mechanism
      seems a little bit uncertain. This is one of the cool things about this paper, but also makes this paragarph
    a bit delicate.}
    This
    result is compatible with the field-based studies, which have evidenced
    increased glacier erosion and pro-glacial sedimentation during deglaciation
    periods \citep[e.g.][]{Koppes.Montgomery.2009,Micheletti.Lane.2016,Lane.etal.2017}, including the ongoing glacier retreat
    \citep[e.g.][]{Bendixen.etal.2017}. Yet this increase has often been attributed
    to higher meltwater availability, promoting sediment transport \citep{Delaney.Adhikari.2020},
    enhanced glacier sliding \citep[][This citation might a bit strange]{Herman.etal.2011}, and hydrofracturing
    (plucking) \citep[][This citation is more about monthly to diurnal variability]{Ugelvig.etal.2018}.

    But importantly, these hydrological processes are not included in our
    model, so that deglaciation surface meltwater is assumed to instantly exit
    the glacier with no effect on glacier dynamics \ian{ \citep[e.g][]{Werder.etal.2013}
      or glacier erosion  \citep[][]{Iverson.etal.2012,Ugelvig.etal.2018}}.
    The enhanced glacier
    velocity during periods of ice retreat is here modelled to result from
    changes of air temperature alone, propagating through diffusion and
    advection into the ice, increasing the glacier's internal energy through
    temperature and englacial water content, thereby facilitating shear strain
    and eventually activating basal sliding \citep[cf.][]{Aschwanden.etal.2012,
    Seguinot.etal.2016}. Observed feedbacks between surface meltwater and basal
    sliding, not included in the model, presumably further accelerated erosion
    during deglaciation \citep{Herman.etal.2011}. \ian{Furthermore, we omit the
      fluvial transport of sediment, which could well increase sediment discharge
      and erosion with increased melt, so long as additional sediment remains
      available \citep{Delaney.Adhikari.2020}. Rapid glacier erosion can be expected
    to accompany the  ongoing 21st-century deglaciation phase. This trend might be
    especially strong on the ice sheets of Greenland and Antarctica where
    polythermal and cold-based glaciers are warming and sliding is increasing
    \citep[e.g.][]{Moon.etal.2012,Overeem.etal.2017}.}

@@

    In many places throughout the Alps, significant modelled cumulative erosion
    occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). Yet
    instantaneous erosion rates show a time-transgressive radial pattern
    (Fig.~\ref{fig:transects}), so that high-altitude rapid erosion is not
    modelled to date from expansive glaciation stages, but instead to stages of
    intermediate glacier cover \citep[Fig.~\ref{fig:hypsogram}; ][]{Barr.etal.2019}.

    For instance, low-elevation piedmont over-deepened troughs such as Lake
    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore, are
    here modelled to be the product of extensive, Last Glacial Maximum-like
    glaciations. \ian{Other modeling work suggests that features such as these
      likely form when water can access the glacier bed and encourage glacier
      sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does not
      include subglacial hydrology, the role of climate warming to concentrating
      glacier erosion to the lower elevations and alpine remains evident.}

    Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valley of the Alps such as Lauterbrunnental
    (Fig.~\ref{fig:hypsogram}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre). While the validity of the model results at high elevation are
    limited by the 1-km resolution and the shallow-ice physics
    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
    of the modelled glacial cycle implies that the numerous Alpine mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
    temperate periods with limited ice-cover such as the present day \ian{
      or over interglacial periods, when glaciers are confined to their cirques
      \citep{Barr.etal.2017,Barr.etal.2019}}.

@@

    Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
    erosion processes, our quantitative results are to be considered with extreme
    uncertainties. Yet to the best of our current knowledge in these respective
    areas of science, we can draw the following qualitative conclusions:
    \begin{itemize}
      \item The non-linear physics of glacier deformation and sliding, combined
        with non-linear empirical erosion laws, results in a very strong
        localization of rapid glacier erosion in regions of fast flow.
      \item Temperature-dependent creep and basal sliding provide a mechanism
        for accelerated erosion during deglaciation periods, irrespective of
        surface meltwater penetration to the glacier bed.
      \item If a non-linear glacier erosion law is used, the climate-induced
        slowdown of erosion counterbalances glacier expansion, so that
        Alpine-wide glacier erosion rates do not correlate with the ice volume.
      \item Rapid glacier erosion is restricted to low elevation during stages
        of glacier advance and maximum glaciation, but propagates up-valley to
        higher elevations during periods of glacier retreat.
      \item The diversity of the Alpine glacial erosion landscape appears to
        be the time-transgressive signature of a variety of glacial stages
        ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
      \item \ian{The high spatialtemporal variability  in erosion suggests that
          the relationship between climate and glacier erosion remains greatly complex.
          As a result, a highly variable response of glacier erosion to climate forcings should be expected.}
    \end{itemize}
